---
title: "Auditing a count matrix before differential expression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Auditing a count matrix before differential expression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

DESeq2, edgeR and limma-voom assume raw counts and a design that can estimate
the effect being tested. They check very little of either. A matrix of CPM
values rounded to whole numbers passes DESeq2 with a routine message and edgeR
and limma in silence; on the airway dataset, that input returns 877
differentially expressed genes where the raw counts give 3,993. `attest` checks
those preconditions before the analysis runs.

It does not analyse expression. Each check adjudicates one claim and reports a
verdict, the evidence for it, and what the inputs could not answer.

```{r}
library(attest)
fx <- readRDS(system.file("extdata", "fixtures.rds", package = "attest"))
m  <- fx$airway            # 3,000 genes x 8 samples, raw counts
```

## One call

```{r}
md <- data.frame(cell = factor(rep(c("N61311", "N052611", "N080611", "N061011"), each = 2)),
                 dex  = factor(rep(c("untrt", "trt"), times = 4), levels = c("untrt", "trt")))
attest(m, metadata = md, design = ~ cell + dex)
```

The overall verdict comes from the checks that look for faults. Detectability is
printed separately: it says how large a change each gene could have shown, which
is a statement about the size of the experiment, not a defect in it.

With a `DESeqDataSet`, the sample sheet and design come from the object, so
`attest(dds)` is the whole call. A check that has nothing to work on is named
under "not run" rather than left out:

```{r}
attest(m)
```

## A matrix that is not counts

```{r}
cpm <- round(t(t(m) / colSums(m)) * 1e6)
attest_counts(cpm)
```

## Rows that are not genes

htseq-count and STAR write their totals into the count table. Left in, they are
analysed as genes:

```{r}
attest_identifiers(rbind(m, "__no_feature" = round(colSums(m) * 0.1)))
```

## A column the design leaves out

Two libraries per donor, analysed as four independent replicates per group:

```{r}
cond  <- factor(rep(c("ctrl", "trt"), each = 4))
donor <- data.frame(cond = cond, donor = rep(paste0("d", 1:4), each = 2))
attest_design(m, donor, ~ cond)
```

## Counts still in a file

```{r}
p <- tempfile(fileext = ".txt")
out <- data.frame(Geneid = rownames(m)[1:500], Chr = "1", Start = 1:500, End = 501:1000,
                  Strand = "+", Length = 1000L, m[1:500, ], check.names = FALSE)
writeLines(c("# Program:featureCounts", paste(names(out), collapse = "\t")), p)
utils::write.table(out, p, sep = "\t", quote = FALSE, row.names = FALSE,
                   col.names = FALSE, append = TRUE)
attest_file(p)$checks$input
```

`Start`, `End` and `Length` are numeric. Built into a matrix by hand, they are
three extra "libraries", and every check on that matrix still passes; the
column names in the file are the only evidence.

## Machine-readable output

```{r}
str(attest_as_list(attest(m))$checks[["value scale"]], max.level = 1)
```
